================================================================================================================
title: "Analysis of personal movement using data from a personal activity monitoring device"
author: "Ezgi"
date: "2 February 2016"
output: html_document
================================================================================================================
## Project Background

It is now possible to collect a large amount of data about personal movement using activity monitoring devices
such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement
- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find 
patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because 
the raw data are hard to obtain and there is a lack of statistical methods and software for processing and 
interpreting the data.

## Data Used in Analysis
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 
minute intervals through out the day. The data consists of two months of data from an anonymous individual 
collected during the months of October and November, 2012 and include the number of steps taken in 5 minute 
intervals each day

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this
dataset.

## Loading and preprocessing the data

1.The code that is needed to load the data:

```{r, echo=TRUE}
install.packages("dplyr")
library(dplyr)
activity_monitoring <- read.csv("./activity.csv")
```

Some statistics to understand data: First 6 rows of table, names of the fields, number of rows and columns and 
dates in the table.

Here is the code:
```{r, echo=TRUE}

head(activity_monitoring)
names(activity_monitoring)
dim(activity_monitoring)

table(activity_monitoring$date)
```

2.Process/transform the data (if necessary) into a format suitable for your analysis:

A check to see how many missing values in all fields is made. 
```{r, echo=TRUE}
steps_na <- which(is.na(activity_monitoring$steps))
steps_na2 <- length(steps_na)

date_na <- which(is.na(activity_monitoring$date))
date_na2 <-length(date_na)

interval_na <- which(is.na(activity_monitoring$interval))
interval_na2 <- length(interval_na)
```

There are `r steps_na2` missing values in steps field, `r date_na2` missing values in date field and `r interval_na2` 
in interval field.


## What is mean total number of steps taken per day?

Missing values in the dataset were ignored for this part of the assignment.

1.The code to calculate the total number of steps taken per day:

```{r, echo=TRUE}

activity_monitoring2 <- activity_monitoring
by_day <- group_by(activity_monitoring2, date)
day_sum <- summarize(by_day, total_steps_perday = sum(steps, na.rm = TRUE))

```

2.Histogram of the total number of steps taken each day

The code and the histogram:

```{r, echo=TRUE}

hist(day_sum$total_steps_perday, col = "red",main = "Total Number of Steps Taken Each Day",
                                              xlab ="Total Number of Steps Taken Each Day")

```

3. The mean and median of the total number of steps taken per day

The code to calculate the mean and median of the total number of steps taken per day:

```{r, echo=TRUE}

median_step_per_day <- median(day_sum$total_steps_perday)
average_step_per_day <- mean(day_sum$total_steps_perday)

```

The mean of the total number of steps taken per day is `r average_step_per_day` and median of the total number 
of steps taken per day is `r median_step_per_day`.


## What is the average daily activity pattern?

1.Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across 
all days (y-axis)

The code and time series plot:

```{r, echo=TRUE}

activity_monitoring3 <- activity_monitoring
by_interval <- group_by(activity_monitoring2, interval)
interval_sum <- summarize(by_interval, total_steps_perinterval = sum(steps, na.rm = TRUE))

qplot(interval, total_steps_perinterval,data=interval_sum, color="red")+labs(y="Total steps per interval")+ 
      labs(title="Average Daily Activity Pattern")+ geom_line()

```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}

max_steps_per_interval <- max(interval_sum$total_steps_perinterval)
max_interval <- interval_sum[which(interval_sum$total_steps_perinterval==max_steps_per_interval),1]

```

`r max_interval` 5-minute interval contains the maximum number of steps n average across all the days in the
dataset.

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of
missing days may introduce bias into some calculations or summaries of the data.

1.The total number of missing values in the dataset (i.e. the total number of rows with NAs)

The code to calculate the total number of missing values in the dataset:

```{r, echo=TRUE}

na_values <- activity_monitoring[which(is.na(activity_monitoring$steps)),]
na_values2 <- nrow(na_values)

```

The total number of missing values in the dataset is `r na_values2`.

2.The strategy chosen for filling in all of the missing values in the dataset is the mean for that 5-minute
interval

```{r, echo=TRUE}

interval_avg <- summarize(by_interval, total_steps_perinterval = mean(steps, na.rm = TRUE))

```

3.A new dataset that is equal to the original dataset is created but with the missing data filled in with
the mean for that 5-minute interval.

```{r, echo=TRUE}

activity_monitoring_notna <- left_join(activity_monitoring,interval_avg) 
activity_monitoring_notna2 <- activity_monitoring_notna[which(is.na(activity_monitoring_notna$steps)),]
activity_monitoring_notna3 <- activity_monitoring_notna[which(!is.na(activity_monitoring_notna$steps)),] 

activity_monitoring_notna4 <- mutate (activity_monitoring_notna2,steps2= total_steps_perinterval)
activity_monitoring_notna5 <- select(activity_monitoring_notna4, date,interval,steps2)

activity_monitoring_notna6 <- select(activity_monitoring_notna3, date,interval,steps2=steps)

activity_monitoring_notna7 <- bind_rows(activity_monitoring_notna5,activity_monitoring_notna6)

```

4.The histogram of the total number of steps taken each day and the mean and median total number of steps taken per day
is calculated.

```{r, echo=TRUE}

activity_monitoring_notna8 <- activity_monitoring_notna7
by_day2 <- group_by(activity_monitoring_notna8, date)
day_sum2 <- summarize(by_day2, total_steps_perday = sum(steps2))

hist(day_sum2$total_steps_perday, col = "red",main = "Total Number of Steps Taken Each Day",xlab ="Total Number of Steps 
     Taken Each Day")

median_step_per_day2 <- median(day_sum2$total_steps_perday)
average_step_per_day2 <- mean(day_sum2$total_steps_perday)

```

The mean of total number of steps taken per day is `r average_step_per_day2` and median is  `r median_step_per_day2`
with the missing data filled in .

These values differ from the estimates from the first part of the assignment. Both median and mean are higher than those
from the data with missing values. Also in the data without missing values the mean and median have the same value.


## Are there differences in activity patterns between weekdays and weekends?

1. A new factor variable is created in the dataset with two levels - "weekday" and "weekend" indicating whether a given 
date is a weekday or weekend day.

The code is below:
```{r, echo=TRUE}

activity_monitoring_notna7$date2 <-as.Date(levels(activity_monitoring_notna7$date), format = "%Y-%m-%d")
activity_monitoring_notna9 <- mutate(activity_monitoring_notna7,if_weekend = 
                                     factor(weekdays(date2) =="Sunday"| weekdays(date2) =="Saturday", 
                                            labels = c("weekday", "weekend")))

```


2.A panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all weekday days or weekend days (y-axis) is made. 

First the code and then the graph:
```{r, echo=TRUE}

install.packages("lattice")
library(lattice)
xyplot(interval ~ steps2 | if_weekend, data = activity_monitoring_notna9, layout = c(5, 1))

```




